---
title: "CoReSh compendium search and derived gene sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CoReSh compendium search and derived gene sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bulkiRNA)
```

This guide remains a vignette because `browseVignettes("bulkiRNA")` is where a
user looks for the order of operations. Its CoReSh chunks are intentionally not
evaluated during package builds. A search needs the roughly 20 GB CoReSh chunk
tree, the optional `qs2` reader, and organism annotation packages. None is
bundled, and `qs2` is absent from the current development image. No output is
fabricated here.

The shared reference tier is expected at
`$REFCACHE_ROOT/coresh/current/preprocessed_chunks/{hsa,mmu}`. The resolved
snapshot tag is attached to results as provenance, so a moving `current`
symlink cannot change bytes silently.

## 1.x: preflight and define queries

CoReSh queries are named integer vectors of Entrez IDs, with at least three
unique IDs each. Keep the exact query definitions in project configuration:
database provenance records their names and sizes, not a copy of every ID.

```{r preflight, eval=FALSE}
Sys.setenv(REFCACHE_ROOT = "/path/to/refcache")

coresh_validate(species = "human")

queries <- list(
  iron_uptake = c(7037L, 4891L, 55240L),
  heme = c(3145L, 3146L, 2035L, 2235L)
)
```

`coresh_validate()` reports all missing prerequisites in one pass. Install
`qs2`, `BiocParallel`, and `org.Hs.eg.db` before continuing; the upstream
`coresh` package itself is not required because its published 0.1.0 release
contains no callable R functions.

## 2.x: search, inspect convergence, and build sets

Each dataset is identified by the `(gse, gpl)` pair. A GSE accession alone is
not unique in the compendium, so keep both columns through every join.

```{r search, eval=FALSE}
hits <- coresh_search(
  queries,
  species = "human",
  n_cores = 4L,
  pvalues = TRUE,
  sample_size = 21L,
  seed = 1L
)

convergent <- coresh_convergence(
  hits,
  top_n = 10L,
  min_queries = 2L
)
convergent
```

With `pvalues = TRUE`, the GESECA p-value is for ranking across the compendium;
it is not a family of independent tests and must not be sent through
`p.adjust()`. `log2err` reports estimator uncertainty, including `Inf` when an
extreme result lies beyond reliable resolution.

The provider below performs the standard composition: search the compendium,
retain top datasets, extract their strongest gene loadings, map them to
symbols, and remove highly redundant sets. It returns the same `gs_db` contract
as every other provider.

```{r provider, eval=FALSE}
coresh_db <- gsdb_coresh(
  queries,
  species = "human",
  top_hits = 5L,
  top_n = 50L,
  min_size = 15L,
  max_size = 500L,
  jaccard_threshold = 0.8,
  n_cores = 4L,
  seed = 1L
)

gsdb_info(coresh_db)$provenance
```

The seed default of 1 is deliberate: it matches the upstream CoReSh
implementation. `bulkirna_stochastic()` records both the seed argument and
what is random.

```{r stochastic}
subset(
  bulkirna_stochastic(quiet = TRUE),
  name %in% c("coresh_search", "gsdb_coresh")
)
```

## 3.x: consume the provider without a special renderer

Once built, CoReSh-derived sets bend no downstream API. A moderated-t rank
vector enters `gs_test()` and the ordinary result renderer consumes the
result. The calls remain unevaluated here because constructing `coresh_db`
requires the external chunk tree.

```{r consume, eval=FALSE}
de_table <- utils::read.csv(
  "03_results/demo/tables/1.6_de_treated_vs_control.csv",
  row.names = 1,
  check.names = FALSE
)
ranks <- gs_ranks(de_table, metric = "t", genes = "Symbol")

res <- gs_test(
  ranks,
  coresh_db,
  contrast = "treated_vs_control",
  min_size = 15L,
  max_size = 500L,
  seed = 123L
)

p <- gs_plot_dot(res, top_n = 15L, sort_by = "padj")
gs_save(p, "03_results/demo/plots/GSEA/2.1_coresh_derived")
```

The result is still a tibble, and `contrast` is still the join key. The
provider owns access to gene sets; compute owns enrichment; the renderer owns
presentation.
