---
title: "Differential expression: counts to a contrast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential expression: counts to a contrast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bulkiRNA)
```

This vignette evaluates a small limma-trend analysis end to end. The counts are
synthetic and are generated in the document; they are not presented as a
biological dataset. Twenty genes are made four-fold higher and twenty are made
four-fold lower in the treated samples, so the fitted contrast has a known
signal.

The house pipeline separates `1.x` ingest and design, `2.x` compute, and `3.x`
render. A contrast is more than a label in a real study: its design and caveats
belong in project configuration. The string label is nevertheless the stable
join key carried into downstream tables.

## 1.x: construct and inspect a `DGEList`

`build_dge()` checks that sample order matches the count matrix and computes
TMM normalization factors. The `set.seed()` call is local to this vignette;
the package's RNG-ownership rule applies to code under `R/`.

```{r synthetic-counts}
set.seed(61)
n_genes <- 400L
n_samples <- 8L
gene_id <- paste0("Gene", seq_len(n_genes))
sample_id <- paste0("Sample", seq_len(n_samples))
group <- factor(rep(c("control", "treated"), each = n_samples / 2L))

base_mean <- stats::rgamma(n_genes, shape = 3, rate = 0.06) + 10
library_factor <- c(0.90, 1.05, 1.10, 0.95, 1.00, 0.92, 1.08, 1.04)
fold_change <- matrix(1, nrow = n_genes, ncol = n_samples)
up_genes <- gene_id[1:20]
down_genes <- gene_id[21:40]
fold_change[1:20, group == "treated"] <- 4
fold_change[21:40, group == "treated"] <- 0.25

mu <- outer(base_mean, library_factor) * fold_change
counts <- matrix(
  stats::rnbinom(length(mu), mu = as.vector(mu), size = 25),
  nrow = n_genes,
  dimnames = list(gene_id, sample_id)
)
samples <- data.frame(group = group, row.names = sample_id)
genes <- data.frame(symbol = gene_id, row.names = gene_id)

dge <- build_dge(counts, samples, genes)
dge$samples[, c("group", "lib.size", "norm.factors")]
```

PCA is a renderer over the normalized count object. It does not fit the DE
contrast.

```{r pca, fig.width=6, fig.height=5}
de_pca(dge, colour_by = "group", title = "Synthetic count structure")
```

## 2.x: fit the moderated contrast

The rank metric used by the gene-set layer is the moderated t-statistic, not
raw fold change. This example uses limma-trend on log-CPM, one of the two house
DE inputs (the other is edgeR quasi-likelihood).

```{r fit}
design <- stats::model.matrix(~ 0 + group, data = dge$samples)
colnames(design) <- levels(group)
log_cpm <- edgeR::cpm(dge, log = TRUE, prior.count = 2)

fit <- limma::lmFit(log_cpm, design)
contrast_matrix <- limma::makeContrasts(
  treated - control,
  levels = design
)
fit <- limma::contrasts.fit(fit, contrast_matrix)
fit <- limma::eBayes(fit, trend = TRUE)
de <- limma::topTable(fit, coef = 1L, n = Inf, sort.by = "none")

de[c(up_genes[1:3], down_genes[1:3]),
   c("logFC", "t", "P.Value", "adj.P.Val")]
```

The fitted table is deliberately left long: one row per gene for this
contrast. In a multi-contrast project, add the contrast string and stack the
rows rather than creating a different schema for each comparison.

```{r rank-and-check}
ranks <- gs_ranks(de, metric = "t")
head(ranks)

planted <- data.frame(
  gene = c(up_genes, down_genes),
  expected = rep(c("up", "down"), each = 20L),
  logFC = de[c(up_genes, down_genes), "logFC"],
  row.names = NULL
)
aggregate(logFC ~ expected, planted, median)
```

## 3.x: render without changing the fit

`de_volcano()` uses raw p-values for vertical resolution but FDR for the
significance decision. The dashed p-value boundary therefore coincides with
the set of FDR-significant genes rather than pretending that raw p = 0.05 is
an FDR cutoff.

```{r volcano, fig.width=7, fig.height=5}
de_volcano(
  de,
  decision_by = "fdr",
  p_cutoff = 0.05,
  fc_cutoff = 1,
  top_n = 4L,
  highlight_gene = c(up_genes[1], down_genes[1]),
  annotate_counts = TRUE,
  title = "Treated vs control (synthetic counts)"
)
```

In a project this naturally becomes `1.6_de.R` for the fit and
`3.0_volcano_viz.R` for the renderer, with the DE table written to
`03_results/<subset>/tables/`. The separation makes it possible to redraw a
figure without refitting the contrast.
