---
title: "GATOM metabolic-network modules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GATOM metabolic-network modules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bulkiRNA)
```

This guide is a vignette so the ordered workflow is discoverable through
`browseVignettes("bulkiRNA")`. Its analysis chunks are intentionally not
evaluated during package builds: GATOM needs the optional `gatom` and `mwcsr`
packages plus about 24 MB of reference files that are not bundled. Fetching
those files would violate an offline, no-refcache build. No output is invented
below.

Install the optional packages first:

```text
BiocManager::install("gatom")
install.packages("mwcsr")
```

## 1.x: resolve the reference inputs

The reference download is a one-time, networked preparation step. Keep the
files in project reference storage and pass the directory explicitly on later
runs. `gatom_refs()` returns the atom-transition network, metabolite database,
and organism annotation together; the metabolite database is required even
when no metabolite DE table is supplied.

```{r references, eval=FALSE}
refs_dir <- "00_data/references/gatom"

# Network required only for this one-time fetch.
gatom_download_refs(
  dir = refs_dir,
  species = "Homo sapiens",
  networks = "kegg"
)

# Offline after the files above exist.
refs <- gatom_refs(
  species = "Homo sapiens",
  dir = refs_dir,
  network = "kegg"
)
refs
```

## 2.x: validate DE input, then solve the module

Start from an unfiltered limma-style table with gene identifiers in `Symbol`.
GATOM's BUM score needs **raw** p-values, not FDR, and `baseMean` must be on a
linear scale. For limma output, `2^AveExpr` supplies that scale.

```{r module, eval=FALSE}
de_table <- utils::read.csv(
  "03_results/demo/tables/1.6_de_treated_vs_control.csv",
  row.names = 1,
  check.names = FALSE
)

gatom_input <- gatom_de(
  de_table,
  id = Symbol,
  pval = P.Value,
  log2FC = logFC,
  baseMean = 2^AveExpr
)

module <- gatom_module(
  gatom_input,
  refs,
  k_gene = 50,
  seed = 42L,
  solver = "rnc"
)
module_genes <- gatom_genes(module)
module_genes
```

`k_gene` is the size dial: smaller values tend to produce larger modules. The
standard sensitivity branches are 25, 50, and 75. Scoring and solving are
stochastic; the default seed 42 preserves the historical GATOM result and is
recorded in the module. The package's complete stochastic registry makes that
choice inspectable.

```{r stochastic}
subset(bulkirna_stochastic(quiet = TRUE), name == "gatom_module")
```

With atom topology, genes label graph **edges**, not vertices. Use
`gatom_genes()` rather than reading vertex symbols. Rendering is a separate
stage and does not recompute the module:

```{r render, eval=FALSE}
gatom_save_html(
  module,
  "03_results/demo/plots/GATOM/2.5_kegg_module.html",
  name = "Treated vs control: KEGG metabolic module"
)
```

The resulting stage layout is `2.5_gatom.R` for compute and
`3.5_gatom_viz.R` for presentation. The DE input and the chosen reference
files should be recorded beside the output; a module without either cannot be
reproduced.
