---
title: "Gene sets: provide, test, render"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene sets: provide, test, render}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bulkiRNA)
```

The gene-set workflow has three boundaries: a provider returns a `gs_db`, a
compute function returns a `gs_result`, and a renderer consumes that result.
Compute never plots, and renderers never recompute a statistic. This separation
is what lets a bundled database, a GMT file, and an inline signature enter the
same analysis.

The example below is fully evaluated without a network connection. The gene
sets are real MitoPathways data shipped in `inst/extdata`; the ranked vector is
synthetic and is labelled as such. A strong positive signal is deliberately
planted in one real pathway so the result has structure worth interpreting.

## 1.x: choose a provider

`gsdb_load()` resolves installed data with `system.file()`, so this works from
both a source checkout and an installed package.

```{r bundled-provider}
gsdb_list()[, c("database", "species")]

db <- gsdb_load(
  "mitopathways",
  species = "Homo sapiens",
  min_size = 15L,
  max_size = 150L
)
db
```

Local GMT and GMX files enter through `gsdb_from_file()`. This small temporary
GMT is constructed only to demonstrate the file boundary; unlike `db` above,
it is not bundled reference data.

```{r file-provider}
gmt <- tempfile(fileext = ".gmt")
writeLines(c(
  "OXPHOS_DEMO\tOxidative phosphorylation demo\tNDUFA1\tNDUFA2\tSDHA",
  "TRANSPORT_DEMO\tTransport demo\tSLC25A3\tSLC25A4\tSLC25A5"
), gmt)

file_db <- gsdb_from_file(
  gmt,
  database = "local_demo",
  database_label = "Local demo",
  species = "Homo sapiens"
)
file_db
```

For signatures already held in R, register the named list directly. Stable,
snake-case database keys are important because `database` becomes a join and
filter key in every result.

```{r inline-provider}
inline_db <- gsdb_register(
  list(
    OXPHOS_DEMO = c("NDUFA1", "NDUFA2", "SDHA"),
    TRANSPORT_DEMO = c("SLC25A3", "SLC25A4", "SLC25A5")
  ),
  database = "inline_demo",
  species = "Homo sapiens",
  database_label = "Inline demo"
)
inline_db
```

MSigDB is also a provider, but it can populate an `msigdbr` cache over the
network. Its call is therefore intentionally not evaluated in a vignette that
must build offline.

```{r msigdb-provider, eval=FALSE}
hallmark <- gsdb_msigdb("Homo sapiens", collection = "H")
```

## 2.x: compute one long result table

The next chunk creates the synthetic ranks. Genes from the selected real
pathway receive a planted positive shift; the remaining mitochondrial and
background genes follow the same null distribution.

```{r planted-ranks}
set.seed(19)
target_id <- names(db)[which.max(lengths(db))]
target_genes <- db[[target_id]]
universe <- unique(c(
  unlist(unclass(db), use.names = FALSE),
  paste0("BACKGROUND_", seq_len(1200L))
))

ranks <- stats::setNames(stats::rnorm(length(universe)), universe)
ranks[target_genes] <- ranks[target_genes] + 4
ranks <- sort(ranks, decreasing = TRUE)

res <- gs_test(
  ranks,
  db,
  contrast = "planted_vs_null",
  min_size = 15L,
  max_size = 150L,
  seed = 123L
)

res[res$pathway_id == target_id,
    c("pathway_id", "contrast", "stat", "stat_type", "padj")]
```

`res` is a tibble with class `gs_result`, not an object with an `@result`
slot. Ordinary table operations and `rbind()` therefore work. `contrast` is a
string column and is the join key when several contrasts are stacked.

The downstream master table has exactly 14 required columns. Supplying the
database and tested universe makes the full-set membership reproducible.

```{r master-table}
master <- gs_to_master(res, db = db, universe = names(ranks))
stopifnot(ncol(master) == 14L)
master[master$pathway_id == target_id,
       c("pathway_id", "contrast", "nes", "padj", "direction")]
```

## 3.x: render, then persist beside the source table

The renderer reads the statistic type from the result, so the axis is labelled
`NES` here and would be labelled differently for another method.

```{r render, fig.width=7, fig.height=5}
p <- gs_plot_dot(
  res,
  top_n = 10L,
  sort_by = "padj",
  title = "Planted enrichment in shipped MitoPathways sets"
)
p
```

In a project, compute and render stages share a number but remain separate.
`gs_save()` enforces the house output rule by writing the figure and its tidy
source table at the same stem:

```{r save, eval=FALSE}
gs_save(
  p,
  "03_results/demo/plots/GSEA/2.1_mitopathways",
  width = 8,
  height = 6
)
```

Seed defaults are not cosmetic: each preserves previously published numbers.
Use `bulkirna_stochastic()` to see which functions consume randomness and why.

```{r stochastic-registry}
bulkirna_stochastic(quiet = TRUE)
```
